PACKER: An Exemplar Model of Category Generation

### 2017-01-21 Manuscripts/cogsci-2017/paper.tex
updates to figs / manuscript

@@

One of the most intriguing capabilities of human cognition is the ability to creatively generate new ideas and concepts. % Some statement about how often people use concepts creatively, prevalence, etc
The creative use of knowledge has, however, scarcely been the subject of scientific inquiry in the field of categorization. Most research on concepts and categorization has conformed to a traditional artificial classification learning paradigm \citep{kurtz2015human}, wherein the primary use of category knowledge is to discriminate between two or more experimenter-defined classes.

@@

\citet{jern2013probabilistic} reported several experiments in support of their model's predictions. In their paradigm, participants are first exposed to members from a set of known categories within a $k$-dimensional feature space (in their experiments, $k=3$). After observing the assigned examples, participants are asked to generate exemplars from a new category within the domain. Participants presented with $k$ scales that can be used to adjust the feature values of the generated stimulus, and are given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that people generated categories possessing the same feature correlations as the known categories in the domain.

@@

% Is this evident? "we are not suggesting Jern \& Kemp have failed to capture a core aspect of generation, but that there are other properties that can be explained through one of the field's oldest traditions."

The PACKER model is in many senses an extension of the highly influential Generalized Context Model of category learning (GCM; \citealp{nosofsky1984choice}). The model assumes that each category is represented by a collection of exemplars within in a $k$-dimensional stimulus space, and that generation is constrained by both similarity to members of the category being generated as well as similarity to members of other categories.

@@

When prompted to generate a new example, the model considers both the summed similarity to examples from other categories as well as the summed similarity to examples in the target category. More formally, the summed similarity $ss$ between generation candidate $y$ and the model's stored exemplars $x_j$ can be computed as:
\begin{equation}
  ss(y, x) = \sum_j{\hat{f_j} s(y, x_j)}
\end{equation}
where $\hat{f_j}$ is a function specifying each example's degree of contribution toward generation. Although $\hat{f_j}$ may be set arbitrarily, in PACKER it is set according to class assignment. For known members of contrast categories, $\hat{f_j} = \phi$. For known members of the target category, $\hat{f_j} = \gamma$. $\phi$ and $\gamma$ are free parameters ($-\infty \leq \phi, \gamma \leq \infty$) controlling the contribution of target- and contrast-category similarity to generation. Larger absolute values for either parameter produce greater consideration of that type of similarity, with values of 0 producing no effect. Negative values produce a `repelling' effect (exemplars are less likely to be generated nearby). Conversely, positive values produce a `pulling' effect (exemplars are more likely to be generated nearby).

@@

@@

\subsubsection{Summary.}
The PACKER model described above explains exemplar and category generation through exemplar similarity (one of the field's oldest traditions). The model proposes that people generate categories by maximizing within-category similarity and minimizing between-category similarity. Unlike the model proposed by \citet{jern2013probabilistic}, the distributional properties of known categories affects performance only insofar as it constrains inter-exemplar similarities. Instead, the \textit{location} of known categories in the domain is crucial because it constrains remaining possible locations for new categories. See Figure \ref{fig:example-prob-spaces} for a depiction of the model's behavior.

@@

The behavioral experiment described below was designed to test a core prediction of the PACKER model: that the \textit{location} of contrast categories (as opposed to their structure), influences generation. The experiment follows the paradigm set by \citet{jern2013probabilistic}: first, participants are exposed to members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B').

\begin{figure}
    \begin{center}
    \input{figs/middle-bottom-conditions.pgf}
    \caption{Experiment conditions.}
    \label{fig:middle-bottom-conditions}
    \end{center}
\end{figure}

Because the goal of the study was to determine if the structure of known categories is the only determinant of generation, our experiment manipulated only the location of the Alpha category within the domain (the structure of the category was held constant). See Figure \ref{fig:middle-bottom-conditions} for a depiction of the experimental conditions. In both conditions, members of the Alpha category are tightly clustered, with equal variance on both features and no correlation between features. Our manipulation is fairly `weak' --  the conditions have only a slight difference in the Y-Axis position of the Alpha category: in the `Middle' condition the Alpha category is placed in the center of the space, in the `Bottom' condition the Alpha category is placed in the bottom-center of the space.

By consequence, the \citet{jern2013probabilistic} model predicts that there should be no difference in the generated categories, as the distributional characteristics of the domain do not differ between the two conditions. However, the PACKER model can predict strong between-condition differences. Assuming exemplars are generated such that each category has strong within-class similarity and weak between-class similarity, top areas of the stimulus space should be used for generation more frequently in the Bottom condition (compared to Middle), as  stimuli in this area possess weaker similarity to members of the Alpha category. Conversely, the bottom areas of the space should be less frequently used for generation in the Bottom condition (compared to Middle), as stimuli in this area possess greater similarity to the Alpha category.

@@

Because the conditions differ only in their location along the Y (vertical) axis (see Figure \ref{fig:middle-bottom-conditions}), our main interest is in generation of Beta examples above and below the contrast category.

Table \ref{table:subset-table} contains the number of participants in each condition who generated at least one example possessing one of the

Fisher's Exact Tests reveal that a greater number of participants assigned to the Middle condition generated at least one Beta example in the bottom rows of the space, $p < 0.001$. Similarly, a greater number of Middle participants generated all four Beta examples in the bottom rows of the space, $p = 0.003$. Participants in the Bottom condition were more marginally more likely to generate at least one Beta example in the top rows of the space, $p = 0.088$, and they were significantly more likely to generate all four examples in the top rows of the space, $p = 0.017$. The conditions did not differ in the number of participants who placed at least one Beta example in the top rows \textit{and} at least one Beta example in the bottom rows, $p = 0.14$.

\begin{table}[!ht]
\begin{center} 
\caption{Behavioral results.} 
\label{table:subset-table} 
\vskip 0.12in
\begin{tabular}{ l r r}
    \textbf{Middle}         & Used top rows & No top rows \\
    \hline
    Used bottom rows        &  31 & 18  \\
    No bottom rows          &  11 &  1  \\
    \\
    \textbf{Bottom}         & Used top rows & No top rows \\
    \hline
    Used bottom rows        & 22 & 8 \\
    No bottom rows          & 29 & 2 \\
\end{tabular}
\end{center} 
\end{table}

@@

@@

\section{Simulations}
\section{Discussion}
\section{Acknowledgments}
Support for this research was provided by the Office of the Vice Chancellor for Research and Graduate Education at the University of Wisconsin - Madison with funding from the Wisconsin Alumni Research Foundation. We would like to thanks Kenneth J. Kurtz for helpful comments on this work.
% any other thanks?

>>>>

%!TEX output_directory = latex_out/

@@

One of the most intriguing capabilities of human cognition is the ability to creatively generate new ideas and concepts. % Some statement about how often people use concepts creatively, prevalence, etc
The creative use of knowledge has, however, infrequently been the subject of scientific inquiry in the field of categorization. Most research on concepts and categorization has instead conformed to a traditional artificial classification learning paradigm \citep{kurtz2015human}, wherein the primary use of category knowledge is to discriminate between two or more experimenter-defined classes.

@@

\citet{jern2013probabilistic} reported several experiments in support of their model's predictions. In their paradigm, participants are first exposed to members from a set of known categories within a $k$-dimensional feature space (in their experiments, $k=3$). After observing the assigned examples, participants are asked to generate exemplars from a new category within the domain. Participants presented with a set of scales scales that can be used to adjust the feature values of the generated stimulus, and are given unlimited time to create each example. As in the classic \cite{ward1994structured} experiment, \citet{jern2013probabilistic} found that people generated categories possessing the same feature correlations as the known categories in the domain.

@@

% Is this evident?: "we are not suggesting Jern \& Kemp have failed to capture a core aspect of generation, but that there are other properties that can be explained through one of the field's oldest traditions."

The PACKER model is an extension of the highly influential Generalized Context Model of category learning (GCM; \citealp{nosofsky1984choice}). The model assumes that each category is represented by a collection of exemplars within in a $k$-dimensional psychological space, and that generation is constrained by both similarity to members of the target category (the category in which a stimulus is being generated) as well as similarity to members of other categories.

@@

When prompted to generate a new example, the model considers both the summed similarity to examples from other categories as well as the summed similarity to examples in the target category. More formally, the summed similarity $ss$ between generation candidate $y$ and the model's stored exemplars $x$ can be computed as:
\begin{equation}
  ss(y, x) = \sum_j{\hat{f_j} s(y, x_j)}
\end{equation}
where $\hat{f_j}$ is a function specifying each example's degree of contribution toward generation. Although $\hat{f_j}$ may be set arbitrarily, in PACKER it is set according to class assignment. For known members of contrast categories, $\hat{f_j} = \phi$. For known members of the target category, $\hat{f_j} = \gamma$. $\phi$ and $\gamma$ are free parameters ($-\infty \leq \phi, \gamma \leq \infty$) controlling the contribution of target- and contrast-category similarity to generation. Larger absolute values for either parameter produce greater consideration of that type of similarity, with values of 0 producing no effect. Negative values produce a `repelling' effect (exemplars are less likely to be generated nearby). Conversely, positive values produce a `pulling' effect (exemplars are more likely to be generated nearby).

@@

\subsubsection{Summary.}

The PACKER model described above explains exemplar and category generation under the traditional exemplar-similarity approach that has guided much of the field's theoretical and empirical work for over three decades. The model proposes that people generate categories by maximizing within-category similarity and minimizing between-category similarity. See Figure \ref{fig:example-prob-spaces} for a depiction of the model's behavior.

@@

Although PACKER appeals to the well-studied exemplar framework, it remains unique as an account of conceptual generation because, unlike the accounts put forward by \citet{jern2013probabilistic} and Ward \& colleagues, it specifies precisely how generated categories should \textit{differ} from known classes in the domain. In this paper, we explore whether an exemplar-similarity approach can offer insight into the creative use of conceptual knowledge. In particular, we focus on a key difference between PACKER and previous accounts: unlike the model proposed by \citeauthor{jern2013probabilistic}, the distributional properties of known categories affects performance in PACKER only insofar as it constrains inter-exemplar similarities. Instead, the \textit{location} of known categories in the domain is crucial because it constrains remaining possible locations for new categories.

@@

The behavioral experiment described below was designed to test the prediction made that the \textit{location} of contrast categories (as opposed to their structure), influences generation. As such, our experiment manipulates only the location of the known categories within the domain (the structure of the categories is held constant).

The experiment follows the paradigm set by \citet{jern2013probabilistic}: first, participants are exposed to members of a known category (`Alpha', or `A'), and are then asked to generate exemplars belonging to a new category (`Beta', or `B'). In both conditions, members of the Alpha category are tightly clustered, with equal variance on both features and no correlation between features. See Figure \ref{fig:middle-bottom-conditions} for a depiction of the experimental conditions. Our manipulation is fairly `weak': the conditions have only a slight difference in the Y-Axis position of the Alpha category. In the `Middle' condition the Alpha category is placed in the center of the space, in the `Bottom' condition the Alpha category is placed in the bottom-center of the space.

\begin{figure}
    \begin{center}
    \input{figs/middle-bottom-conditions.pgf}
    \caption{Conditions tested in the behavioral experiment.}
    \label{fig:middle-bottom-conditions}
    \end{center}
\end{figure}

Because the distributional characteristics of the domain do not differ between the Middle and Bottom conditions, the \citet{jern2013probabilistic} model predicts that there should be no difference in the categories generated within each domain. However, the PACKER model proposes that exemplars are generated such that each category has strong within-class similarity and weak between-class similarity. By consequence, the upper areas of the stimulus space should be used for generation more frequently in the Bottom condition compared to Middle, as stimuli in this area possess weaker similarity to members of the Alpha category. Conversely, the bottom areas of the space should be less frequently used for generation in the Bottom condition compared to Middle, as stimuli in this area possess greater similarity to the Alpha category.

@@

Because the conditions differ only in their location along the vertical axis (see Figure \ref{fig:middle-bottom-conditions}), our main interest is in the generation of Beta examples above and below the contrast category. A highly detailed picture of these data is shown in Figure \ref{fig:middle-bottom-yranges}, wherein each participant's generated category is depicted in terms of its use of the vertical axis. As is evident in Figure \ref{fig:middle-bottom-yranges}, we observed a great deal of variance in generation strategy: whereas some participants generated all four Beta examples within a narrow range of the Y axis, others generated Beta examples along a fairly wide range.

@@

To obtain a more abstracted view of our results, we determined the number of participants in each condition who placed at least one Beta example within the top `rows' of the space (one of the top two possible Y axis values), as well as the number of participants who placed at least one example in one of the bottom rows. The resulting contingency tables data are shown in Table \ref{table:subset-table}.

\begin{table}
\begin{center} 
\caption{Behavioral results.} 
\label{table:subset-table} 
\vskip 0.12in
\begin{tabular}{ l r r}
    \textbf{Middle}         & Used top rows & No top rows \\
    \hline
    Used bottom rows        &  31 & 18  \\
    No bottom rows          &  11 &  1  \\
    \\
    \textbf{Bottom}         & Used top rows & No top rows \\
    \hline
    Used bottom rows        & 22 & 8 \\
    No bottom rows          & 29 & 2 \\
\end{tabular}
\end{center} 
\end{table}

Fisher's Exact Tests reveal that a greater number of participants assigned to the Middle condition generated at least one Beta example in the bottom rows of the space, $p < 0.001$. Participants in the Bottom condition were more marginally more likely to generate at least one Beta example in the top rows of the space, $p = 0.088$. The conditions did not differ in the number of participants who placed at least one Beta example in the top rows \textit{and} at least one Beta example in the bottom rows, $p = 0.14$. Finally, a greater number of Middle participants generated all four Beta examples in the bottom rows of the space, $p = 0.003$, and a greater number of Bottom participants generated all four examples in the top rows of the space, $p = 0.017$.

@@

\section{Simulations}
\section{Discussion}
\section{Acknowledgments}
Support for this research was provided by the Office of the Vice Chancellor for Research and Graduate Education at the University of Wisconsin - Madison with funding from the Wisconsin Alumni Research Foundation. We would like to thank Kenneth J. Kurtz for helpful comments on this work.
% any other thanks?
